---
title: "Unit 4 - Data Wrangling"
subtitle: "PSYC 640 - Fall 2024"
author: "Dustin Haraden, PhD"
format: 
  revealjs:
    multiplex: true
    scrollable: true
    slide-number: true
    incremental: false
    touch: true
    code-overflow: wrap
    theme: dark
execute: 
  echo: true
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r, include = F}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Reminders

::: nonincremental
-   Journal Entries

-   Reverse Results - Due 9/17
:::

------------------------------------------------------------------------

## Last Class

::: nonincremental
-   Entry into the tidyverse with dplyr
:::

------------------------------------------------------------------------

## Today...

Continuing on this trend to get more comfortable with all that `dplyr` can do!

[This is a nice website that walks through a lot of this as well](https://p4a.jhelvy.com/data-wrangling){target="_blank"}

```{r, results = 'hide', message = F, warning = F}
# File management
library(here)
# for dplyr, ggplot2
library(tidyverse)
#Loading data
library(rio)

#Remove Scientific Notation 
options(scipen=999)
```

------------------------------------------------------------------------

## But first...

We need to revisit how to read the [ggplot2 cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf){target="_blank"}!

A lot of cheatsheets will have something that looks like this:

```{r}
c <- ggplot(mpg, aes(hwy))

## Then later on it will say

c + geom_histogram(binwidth = 5)
```

------------------------------------------------------------------------

## WHAT IS `c`?!

`c` is just an object that is referring to the first line that we have in a ggplot function

```{r eval=FALSE}
c <- ggplot(mpg, aes(hwy))

## So this...
c + geom_histogram(binwidth = 5)

## Is the same as this...
mpg %>% 
  ggplot(aes(hwy)) + 
  geom_histogram(binwidth = 5)
```

------------------------------------------------------------------------

## Back to regular scheduled programming

Let's first start by opening our **`Project`**

Then, create a new Notebook/Markdown Document that we will use for today

Setup the libraries and bring in the data

-   We will use the "Sleep Data" file

```{r}
#import sleep data - Your path may be different
sleep_data <- import(here("lectures", "data", "Sleep_Data.csv"))
```

# select()

![](/images/select().png){fig-align="center"}

------------------------------------------------------------------------

## Variable Renaming

Ooof…These variable names look terrible! How do we update those? Our brains would break having to remember how to connect them and having to reference another document. 

First, let’s get rid of the variables that we don’t need for the moment: 

-   Use `select()` to remove the following variables: 

    ```{r}
    sleep_data <- sleep_data %>% 
      select(-c('StartDate', 'EndDate', 'Status', 'Progress', 
      'Duration__in_seconds_', 'Finished', 'RecordedDate',
      'DistributionChannel', 'UserLanguage', 'SC0', 'SC1',
      'SC2', 'SC3', 'SC4', 'SC5', 'Attention'))
    ```

------------------------------------------------------------------------

Now we can update the names

-   Using the `names()` command, we can rename ***all*** columns in the file 

-   Here is a list so that it makes it easier to do instead of having to type everything out…that would just be cruel

```{r}
names(sleep_data) <- c('age', 'gender', 'roommate', 'other_sleep', 'bed_read', 'bed_study', 'bed_hw', 'attention1', 'bed_internet',
                       'bed_tv', 'bed_eat', 'bed_friends', 'bed_videogame', 'bed_readp', 'BL_1', 'BL_2', 'BL_3', 'BL_4',
                       'BL_5', 'attention2', 'BL_6', 'BL_7', 'sleepsat1', 'sleepsat2', 'sleepsat3', 'sleepsat4', 'sleepsat5',
                       'sleepsat6', 'ESS1', 'ESS2', 'ESS3', 'ESS4', 'ESS5', 'ESS6', 'ESS7', 'ESS8', 'ESS00', 'ashs1', 'ashs2', 
                       'ashs3', 'ashs4', 'ashs5', 'attention3', 'ashs6', 'ashs7', 'ashs8', 'ashs9', 'ashs10', 'ashs11', 
                       'ashs12', 'ashs13', 'ashs14', 'ashs15', 'ashs16', 'ashs17', 'attention4', 'ashs18', 'ashs19', 
                       'ashs20', 'ashs21', 'ashs22', 'ashs23', 'ashs24', 'ashs26', 'ashs27', 'ashs28', 'attention5', 
                       'ashs29', 'ashs30', 'ashs31', 'ashs32', 'ashs33', 'id')
```

# filter()

![](/images/filter().png){fig-align="center"}

------------------------------------------------------------------------

## Extract Rows

The 'filter()' function is used to subset observations based on their values.

The result of filtering is a data frame with the same number of columns as before but fewer rows.

The first argument is data and subsequent arguments are logical expressions that tell you which observations to retain in the data frame.

```{r}
filter(starwars, hair_color == "none" | eye_color == "black")
```

## Filter observations

We can now generate a subset of observations based on a particular value

Since the `age` variable has been giving us some trouble, let's filter based on that variable to only include appropriate ages for "college aged" participants

```{r}
# Assigning it to a new variable
mod_sleep_data <- sleep_data %>% 
  filter(age >= 17, age <= 25)
```

------------------------------------------------------------------------

## Filter Obs. -  Check

Generating a bar chart to see if it worked out!

```{r}
mod_sleep_data %>% 
  ggplot(aes(age)) + 
  geom_bar()
```

# mutate()

![](/images/mutate().png){fig-align="center"}

------------------------------------------------------------------------

## Compute new variable

We often need to make a sum/mean score for a variable of interest

For example, in the `mod_sleep_data`, we have the [Epworth Sleepiness Scale](https://nasemso.org/wp-content/uploads/neuro-epworthsleepscale.pdf){target="_blank"}

Take a look at the scoring of the ESS and compute the total score labelled `ess_total`

------------------------------------------------------------------------

## `ess_total`

It is a sum score with a scale of 0 - 24

```{r}
# We will assign it to the same dataset
mod_sleep_data <- mod_sleep_data %>% 
  mutate(ess_total = ESS1 + ESS2 + ESS3 + 
           ESS4 + ESS5 + ESS6 + ESS7 + ESS8)
```

------------------------------------------------------------------------

## Visualize

```{r}
mod_sleep_data %>% 
  ggplot(aes(ess_total)) + 
  geom_bar()
```

::: incremental
-   But why are there some above 24??
:::

------------------------------------------------------------------------

## ESS Variables

The original items are supposed to be on a 0-3 scale

The data that we have are on a 1-4 scale

***What should we do?!***

------------------------------------------------------------------------

## ESS Rescale

We could subtract 1 from each item before creating another sum

```{r}
mod_sleep_data <- mod_sleep_data %>% 
  mutate(ESS1m1 = ESS1 - 1, 
         ESS2m1 = ESS2 - 1,
         ESS3m1 = ESS3 - 1,
         ESS4m1 = ESS4 - 1,
         ESS5m1 = ESS5 - 1,
         ESS6m1 = ESS6 - 1,
         ESS7m1 = ESS7 - 1, 
         ESS8m1 = ESS8 - 1) %>% 
  mutate(ess_total2 = ESS1m1 + ESS2m1 + ESS3m1 + 
           ESS4m1 + ESS5m1 + ESS6m1 + ESS7m1 + ESS8m1)
```

------------------------------------------------------------------------

## ESS Rescale

Or we could combine all the `1`'s that we are subtracting and just take that from the total score

```{r}
mod_sleep_data <- mod_sleep_data %>% 
  mutate(ess_total = ESS1 + ESS2 + ESS3 + 
           ESS4 + ESS5 + ESS6 + ESS7 + ESS8) %>% 
  mutate(ess_total3 = ess_total-8)
```

------------------------------------------------------------------------

## ESS Rescale

Let's see what the differences are between these methods

```{r}
mod_sleep_data %>% 
  select(contains("_total")) %>% 
  head()
```

------------------------------------------------------------------------

## Rescaling Conclusion

It is good to always know what the original scaling needs to be (e.g., 1-4 or 0-3)

Extremely important to visualize the data! ALWAYS VISUALIZE

There is also another way to allow for the `sum()` function in `mutate()`, but we'll get to that later
