---
title: "Week 4 - The Inference Machine"
author: "YOUR NAME + PARTNER'S NAME"
format: docx
execute:
  error: true   # lets the file render even before every blank is filled
---

```{r setup}
#| message: false
library(tidyverse)
```

# Part 0: Design warm-up

> "Grad students who work more than 50 hours a week are more burned out, so universities should cap lab hours."

1.  Predictor / outcome and how they might be operationalized:

2.  Association or causal claim?

3.  Which rules of causality can a survey satisfy, and which can't it?

4.  One possible confound:

5.  An experiment (or why not):

# Part 1: Build a population

```{r population}
set.seed(640)
population <- tibble(
  student_id = 1:10000,
  burnout    = round(rnorm(10000, mean = 50, sd = 12))
)

# True population mean
true_mean <- mean(population$___)
true_mean

# Histogram of burnout
ggplot(population, aes(x = ___)) +
  geom_histogram(binwidth = 2)
```

❓ Why can we almost never calculate the true mean in real research?

**Answer:**

# Part 2: One sample vs. many samples

```{r one-sample}
one_sample <- population %>%
  slice_sample(n = ___)

mean(one_sample$burnout)
```

❓ Why is your mean different from your neighbor's?

**Answer:**

```{r many-samples}
# Draws ONE sample of size n and returns its mean (provided for you)
sample_mean <- function(n) {
  population %>%
    slice_sample(n = n) %>%
    summarize(m = mean(burnout)) %>%
    pull(m)
}

set.seed(2)
sims <- tibble(n = rep(c(10, 50, 200), each = ___)) %>%
  mutate(sample_mean = map_dbl(n, sample_mean))   # this may take ~10 seconds

# Center and spread of the sample means for each n
sims %>%
  group_by(___) %>%
  summarize(avg_of_means = mean(sample_mean),
            se           = sd(sample_mean))

# Compare to the formula for the standard error
12 / sqrt(c(10, 50, 200))
```

```{r sampling-dist-plot}
ggplot(sims, aes(x = sample_mean)) +
  geom_histogram(binwidth = 0.5) +
  geom_vline(xintercept = ___, color = "red") +
  facet_wrap(~ n, ncol = 1, labeller = label_both)
```

❓ What happens to the center and the spread as n increases? Which idea from lecture is this?

**Answer:**

❓ How do the SDs of the sample means compare to `12 / sqrt(n)`?

**Answer:**

# Part 3: The null world

```{r null-world}
# One "study": two groups drawn from the SAME population (no true difference)
null_study <- function(n_per_group) {
  g1 <- sample(population$burnout, n_per_group)
  g2 <- sample(population$burnout, n_per_group)
  t.test(g1, g2)$p.value
}

null_study(20)   # run this line a few times!

set.seed(3)
null_ps <- tibble(p = map_dbl(1:1000, ~ null_study(20)))

# Proportion of "significant" results
null_ps %>%
  summarize(false_positive_rate = mean(p < ___))

ggplot(null_ps, aes(x = p)) +
  geom_histogram(binwidth = 0.05, boundary = 0) +
  geom_vline(xintercept = .05, color = "red")
```

❓ What proportion was significant? What do we call those results?

**Answer:**

❓ What shape is the p-value histogram? Does a single p = .03 "prove" anything?

**Answer:**

# Part 4: A real effect

```{r real-effect}
# Group 2 truly scores `effect` points higher
effect_study <- function(n_per_group, effect = 5) {
  g1 <- sample(population$burnout, n_per_group)
  g2 <- sample(population$burnout, n_per_group) + effect
  t.test(g1, g2)$p.value
}

set.seed(4)
power_sims <- tibble(n_per_group = rep(c(___, ___, ___), each = 1000)) %>%
  mutate(p = map_dbl(n_per_group, effect_study))

# Power = proportion of studies that detected the real effect
power_sims %>%
  group_by(n_per_group) %>%
  summarize(power = mean(___))
```

❓ With 20 per group, how often did we miss the real effect? What is that error called?

**Answer:**

❓ What does this tell you about planning the burnout study's sample size?

**Answer:**

# Challenge: The forking paths

```{r forking-paths}
# Each simulated researcher tests 5 outcomes and keeps the smallest p-value
set.seed(5)
hack <- tibble(sim = 1:1000) %>%
  mutate(min_p = map_dbl(sim, ~ min(map_dbl(1:5, function(i) null_study(20)))))

hack %>%
  summarize(any_significant = mean(min_p < .05))

1 - .95^5
```

❓ How does this connect to the replication crisis?

**Answer:**

# Wrap-up

*"A p-value tells me \_\_\_, but it does not tell me \_\_\_."*

**Answer:**
