---
title: " "
author: "Me"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{r}
# File management
library(here)
# for dplyr, ggplot2
library(tidyverse)
#Loading data
library(rio)

# Pretty variable names
library(janitor)

library(easystats)

#Remove Scientific Notation 
options(scipen=999)

life_exp <- import(here("Data", "Life_Expectancy.csv")) %>% 
  janitor::clean_names()

```

Life expectancy as the outcome or variable that I want to predict
(continuous)

# Main Question: What factors influence life expectancy?

## Sub Question: 

Life expectancy different by developed vs. not

```{r}
t.test(life_exp$life_expectancy ~ life_exp$status)
```

Note: across all years

There is a mean group difference between Developed Countries (m = 79.2,
sd = XX) and Developing Countries (m = 67.11, sd = XX ), such that
individuals in developed countries have a higher life expectancy.

Filtering to only include a single year of data so that there is no
dependence in the variable (one country had multiple observations)

```{r}
life_exp2015 <- life_exp %>% 
  filter(year == 2015)

t.test(life_exp2015$life_expectancy ~ life_exp2015$status)
```

We aren't able to run ANOVA because we don't have a categorical variable
that has more than 2 groups. This all comes back to how we are measuring
the data.

### Does life expectancy change over time

Trying to predict life expectancy by our time variable

Y = b0 + b1X +error

```{r}
reg1 <- lm(life_expectancy ~ year, data = life_exp)

summary(reg1)

life_exp %>% 
  ggplot(aes(year, life_expectancy)) +
  geom_point() + 
  geom_smooth(method = "lm")
```

For every 1 unit increase in x (for every year), life expectancy
(y-variable) will increase by b1 units (0.35).
