---
title: "Week 05: Simple Regression"
subtitle: "Date: September 24, 2026"
footer:  "[course-website](https://dharaden.github.io/psyc640/)"
logo: "images/640_hex.png"
format:
  revealjs:
    theme: [default, brand, clean.scss]
    multiplex: false
    transition: fade
    transition-speed: fast
    slide-number: true
    incremental: false
    chalkboard: true
    overview: false
    scrollable: true
    code-line-numbers: true
editor: visual
execute:
  echo: true
  freeze: auto
---

## Today...

**The Linear Model** — one machine, most of the course

- Why a line? (and why regression *first*)
- Data = Model + Error
- Where the intercept and slope come from
- Reading a slope out loud
- How well does the line fit? ($R^2$, effect size)
- The punchline from Lab 5

```{r, results = 'hide', message = F, warning = F}
# File management
library(here)
# for dplyr, ggplot2
library(tidyverse)
# Loading data
library(rio)
# tidy model output
library(broom)
# descriptives
library(psych)
# pretty tables
library(sjPlot)

#Remove Scientific Notation
options(scipen = 999)

cah <- import(here("files", "data", "CAH.csv"))
```

------------------------------------------------------------------------

## First, a confession about Lab 5

In Lab 5 you ran three things:

1.  A **t-test** comparing two groups
2.  A **correlation** between two continuous variables
3.  An `lm()` — the "sneak peek" 👀

And the numbers came back *the same*.

. . .

That was not a bug, and it was not a coincidence.

::: callout-important
Regression is not a new test that sits next to t-tests and correlations. It is the **general machine** that both of them are special cases of.
:::

------------------------------------------------------------------------

## Why I teach regression first

Regression is a general data analytic system

- Lots of things fall under the umbrella of regression

- It handles many forms of relations and types of variables

- The output gives you **effect sizes** *and* **statistical significance**

- We can add multiple predictors and account for their intercorrelations

. . .

Learn one machine well, then add parts to it. That is the whole plan for the next eight weeks.

------------------------------------------------------------------------

### Uses for regression

::: incremental
- **Adjustment**: Take into account (*control*) known effects in a relationship

- **Prediction**: Develop a model based on what has happened previously to predict what will happen in the future

- **Explanation**: Examine the influence of one or more variables on some outcome
:::

# Part 1: Building the Line

------------------------------------------------------------------------

### The core idea

With regression, we are ***building a model*** that we think best represents the data, and the broader world

$$
Data = Model + error
$$

. . .

Every model we build for the rest of the semester is a variation on this sentence.

------------------------------------------------------------------------

## A line you already know

At the simplest form we draw a line to characterize the linear relationship between two variables, so that for any value of `x` we have an estimate of `y`

$$
Y = mX + b
$$

::::::: columns
:::: {.column width="50%"}
::: incremental
- Y = Outcome Variable (DV)

- m = Slope Term
:::
::::

:::: {.column width="50%"}
::: incremental
- X = Predictor (IV)

- b = Intercept
:::
::::
:::::::

------------------------------------------------------------------------

### Let's "science up" the equation

$$
Y_i = b_0 + b_1X_i + e_i
$$

This captures how we calculate **each observation** ( $Y_i$ )

$$
\hat{Y_i} = b_0 + b_1X_i
$$

This gives the "best guess" or *expected* value of $Y$ given $X$

------------------------------------------------------------------------

### Expected vs. Actual

$$Y_i = b_{0} + b_{1}X_i + e_i \qquad \hat{Y_i} = b_{0} + b_{1}X_i$$

$\hat{Y}$ signifies the **fitted** score — no error. We interpret it as "on average."

The difference between the fitted and the observed score is the **residual**:

$$Y_i - \hat{Y_i} = e_i$$

::: callout-note
There is a different $e$ value for **each observation** in the dataset. The residual is the part of a person your model gets wrong — hold onto that, it becomes the whole story in a few weeks.
:::

------------------------------------------------------------------------

## OLS

::: incremental
- How do we find the regression estimates?

- **O**rdinary **L**east **S**quares estimation

- Minimizes deviations

  - $$ min\sum(Y_{i} - \hat{Y} ) ^{2} $$

- Other estimation procedures are possible (and necessary in some cases)
:::

------------------------------------------------------------------------

## Some data

```{r plot1}
#| code-fold: true

set.seed(142)
x.1 <- rnorm(10, 0, 1)
e.1 <- rnorm(10, 0, 2)
y.1 <- .5 + .55 * x.1 + e.1
d.1 <- data.frame(x.1, y.1)
m.1 <- lm(y.1 ~ x.1, data = d.1)
d1.f <- augment(m.1)

ggplot(d1.f, aes(x = x.1, y = y.1)) +
  geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) +
  theme_bw(base_size = 20)
```

------------------------------------------------------------------------

## The fitted values ( $\hat{Y}$ )

```{r}
#| code-fold: true

ggplot(d1.f, aes(x = x.1, y = y.1)) +
  geom_point(size = 2) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) +
  theme_bw(base_size = 20)
```

------------------------------------------------------------------------

## The residuals ( $e_i$ )

```{r}
#| code-fold: true

ggplot(d1.f, aes(x = x.1, y = y.1)) +
  geom_point(size = 2) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) +
  geom_segment(aes(xend = x.1, yend = .fitted), color = "red") +
  theme_bw(base_size = 20)
```

------------------------------------------------------------------------

## Put it together

```{r, message = F}
#| code-fold: true

ggplot(d1.f, aes(x = x.1, y = y.1)) +
  geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) +
  geom_segment(aes(xend = x.1, yend = .fitted), color = "red") +
  theme_bw(base_size = 20)
```

------------------------------------------------------------------------

## Compare to a bad fit

Which has more \[**RED\]**{style="color: red;"}

::::: columns
::: {.column width="50%"}
```{r, echo=FALSE, message = F}
ggplot(d1.f, aes(x = x.1, y = y.1)) +
  geom_point(size = 2) +
  geom_smooth(method = lm, se = FALSE) +
  geom_point(aes(y = .fitted), shape = 1, size = 2) +
  geom_segment(aes(xend = x.1, yend = .fitted), color = "red") +
  theme_bw(base_size = 20)
```
:::

::: {.column width="50%"}
```{r, echo = F, message = F}
new.i = 1.1
new.slope = -0.7
d1.f$new.fitted = 1.1 - 0.7 * d1.f$x.1

ggplot(d1.f, aes(x = x.1, y = y.1)) +
  geom_point(size = 2) +
  geom_abline(intercept = new.i, slope = new.slope, color = "blue", linewidth = 1) +
  geom_point(aes(y = new.fitted), shape = 1, size = 2) +
  geom_segment(aes(xend = x.1, yend = new.fitted), color = "red") +
  theme_bw(base_size = 20)
```
:::
:::::

------------------------------------------------------------------------

## OLS

The line that yields the smallest sum of squared deviations

$$\Sigma(Y_i - \hat{Y_i})^2$$ $$= \Sigma(Y_i - (b_0+b_{1}X_i))^2$$ $$= \Sigma(e_i)^2$$

. . .

You *could* try many different coefficients until you find the pair with the smallest sum of squared deviations. Luckily, there are simple calculations that yield the OLS solution every time.

------------------------------------------------------------------------

## Where does the slope come from? 🤔

Starting with the **covariance** between two variables:

$$cov_{xy} = \frac{\sum(x-\bar{x})(y-\bar{y})}{N-1}$$

and the correlation, which is just a standardized covariance:

$$r_{xy} = \frac{cov_{xy}}{s_x s_y}$$

. . .

The regression coefficient is built from exactly the same pieces.

------------------------------------------------------------------------

## Regression coefficient, $b_{1}$

$$b_{1} = \frac{cov_{XY}}{s_{x}^{2}} = r_{xy} \frac{s_{y}}{s_{x}}$$

What units is the regression coefficient in?

. . .

The slope equals the estimated change in **Y** for a **1-unit** change in **X**

::: callout-tip
A correlation is unitless. A slope is **in the units of your study**. That is why a slope is often the more useful number to report to a human being.
:::

------------------------------------------------------------------------

## Slope *is* the correlation (rescaled)

$$\Large b_{1} = r_{xy} \frac{s_{y}}{s_{x}}$$

If the standard deviation of both X and Y is equal to 1:

$$\Large b_1 = r_{xy} \frac{s_{y}}{s_{x}} = r_{xy} \frac{1}{1} = r_{xy} = \beta_{yx}$$

. . .

**Standardize both variables and the slope becomes the correlation.** Same information, different clothes.

------------------------------------------------------------------------

## Estimating the intercept, $b_0$

- The intercept adjusts for differences in the means of X and Y

$$\hat{Y_i} = \bar{Y} + r_{xy} \frac{s_{y}}{s_{x}}(X_i-\bar{X})$$

- It is where the regression line crosses the y-axis, at $X = 0$

- It ensures the line runs through the point $(\bar{X}, \bar{Y})$

$$\Large b_0 = \bar{Y} - b_1\bar{X}$$

------------------------------------------------------------------------

## 🔍 Visualizing OLS

Play with these — they make the "least squares" part click:

<https://setosa.io/ev/ordinary-least-squares-regression/>

[https://observablehq.com/\@yizhe-ang/interactive-visualization-of-linear-regression](https://observablehq.com/@yizhe-ang/interactive-visualization-of-linear-regression)

# Part 2: Reading a Line Out Loud

------------------------------------------------------------------------

## Our data today 🃏

**Cards Against Humanity: *Pulse of the Nation*** — a national poll that asked serious demographic questions *and* some very strange ones.

```{r}
cah %>%
  select(age, transformers, books) %>%
  psych::describe()
```

------------------------------------------------------------------------

## 🛠️ Activity Stop 1: One Line, Two Stories

::: callout-tip
**Part A: Follow along** · \~20 min · Pairs
:::

We fit this model together, from the top:

- Plot it, then guess *r* **before** you run it
- `lm()`, find the intercept and slope
- Say the slope in a unit that means something
- Predict a person and compute their residual

[Open the activity](../class-activities/wk5_linear-model.html){.button target="_blank"}

------------------------------------------------------------------------

## Step 1: ALWAYS visualize first

```{r, message = F, warning = F}
#| code-fold: true

ggplot(cah, aes(x = age, y = transformers)) +
  geom_point(alpha = .3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Age (years)",
       y = "Transformers movies seen") +
  theme_bw(base_size = 20)
```

------------------------------------------------------------------------

## By hand, then by R

```{r}
r        <- cor(cah$age, cah$transformers, use = "pairwise")
s_age    <- sd(cah$age, na.rm = TRUE)
s_tf     <- sd(cah$transformers, na.rm = TRUE)
m_age    <- mean(cah$age, na.rm = TRUE)
m_tf     <- mean(cah$transformers, na.rm = TRUE)

b1 <- r * (s_tf / s_age)
b0 <- m_tf - (b1 * m_age)

c(r = r, b1 = b1, b0 = b0)
```

------------------------------------------------------------------------

## In `R`

```{r}
fit.1 <- lm(transformers ~ age, data = cah)
summary(fit.1)
```

::: notes
Point at the Estimate column first, then the t and p. Note the slope's t is identical to cor.test — we prove it on the next slides.
:::

------------------------------------------------------------------------

## wait...they aren't the same? Rounding?

The results are close, but not identical. What's the deal??

::::: columns
::: {.column width="50%"}
```{r}
c(r = r, b1 = b1, b0 = b0)
```
:::

::: {.column width="50%"}
```{r}
tidy(fit.1)
```
:::
:::::

. . .

**MISSING DATA**

Our data has 62 participants with missing values

`lm()` just kicks folks out with `NA` - listwise deletion

Hand calculation used all participants in the data

------------------------------------------------------------------------

## Comparing...again

This time we are going to listwise delete participants.

```{r}
cah_tf <- cah %>% drop_na(age, transformers)
r        <- cor(cah_tf$age, cah_tf$transformers, use = "pairwise")
s_age    <- sd(cah_tf$age, na.rm = TRUE)
s_tf     <- sd(cah_tf$transformers, na.rm = TRUE)
m_age    <- mean(cah_tf$age, na.rm = TRUE)
m_tf     <- mean(cah_tf$transformers, na.rm = TRUE)

b1 <- r * (s_tf / s_age)
b0 <- m_tf - (b1 * m_age)

c(r = r, b1 = b1, b0 = b0)
```

```{r}
tidy(fit.1)
```

------------------------------------------------------------------------

## Saying the slope out loud

$$\widehat{transformers} = 2.58 - 0.025(age)$$

::: incremental
- "Each **additional year** of age is associated with **0.025 fewer** Transformers movies seen."

- That sounds like nothing. So **rescale it to a unit people care about**:

  - 10 years → about **0.25 fewer** movies
  - 40 years → about **one whole movie** fewer

- Note the word **associated**. Nobody was randomly assigned an age.
:::

------------------------------------------------------------------------

## And the intercept?

$b_0 = 2.58$ = the predicted number of Transformers movies for a person aged **zero**.

. . .

::: callout-warning
## Careful

Nobody in this sample is under **18**. A prediction at age 0 is **extrapolation** — the model has never seen data anywhere near there.

The intercept is an **anchor for the line**, not a finding.
:::

. . .

When zero is not meaningful, **centering** the predictor fixes the interpretation (we will do this for real in Week 9).

------------------------------------------------------------------------

## Predict a person

```{r}
# predicted movies for a 20-year-old
b0 + b1 * 20
```

. . .

A 20-year-old in the data who has seen **4** movies:

$$e_i = Y_i - \hat{Y_i} = 4 - 2.07 = +1.93$$

**Positive residual** → above the line → the model **under-predicted** this person.

------------------------------------------------------------------------

## ☕ Break {background-color="#C9ADE0"}

**10 minutes**

When we come back: how do we know if this line is any *good*?

# Part 3: How Well Does the Line Fit?

------------------------------------------------------------------------

## Statistical inference

- The way the world is = our model + error

- How good is our model? Does it "fit" the data well?

. . .

To assess fit, we examine the proportion of variance in our outcome that is "explained" by our model.

To do that, we **partition the variance** into two categories: variability captured by the model, and variability that is not.

------------------------------------------------------------------------

## Partitioning variation

Start with the relationship between observed $Y$ and fitted $\hat{Y}$:

$$Y_i = \hat{Y}_i + e_i$$

. . .

Variances are **additive when two variables are independent**. If C = A + B, then the variance of C is the sum of the variances of A and B.

. . .

Why can we use that rule here?

::: notes
Students must recognize that Y-hat and e are uncorrelated — they must be, by the way OLS is built.
:::

------------------------------------------------------------------------

## Partitioning variation

::::: columns
::: {.column width="50%"}
[$\hat{Y}_i$ and $e_i$ are independent of each other. So the variance of $Y$ equals the variance of $\hat{Y}$ plus the variance of $e$.]{style="font-size:80%"}

$$\large s^2_Y = s^2_{\hat{Y}} + s^2_{e}$$
:::

::: {.column width="50%"}
[Variances are sums of squares divided by N−1, and all have the same sample size here, so:]{style="font-size:80%"}

$$\large SS_Y = SS_{\hat{Y}} + SS_{\text{e}}$$
:::
:::::

. . .

$$ SS_Y = SS_{\text{Model}} + SS_{\text{Residual}}$$

------------------------------------------------------------------------

## Coefficient of Determination

$$\Large \frac{SS_{Model}}{SS_{Y}} = \frac{s_{Model}^2}{s_{y}^2} = R^2$$

$R^2$ is the **proportion of variance in Y explained by the model**

. . .

$$\large \sqrt{R^2} = r_{Y\hat{Y}}$$

With one predictor, that is just our old friend $r$.

------------------------------------------------------------------------

## Ta da! 🎉

```{r}
# the correlation, squared
cor(cah$age, cah$transformers, use = "pairwise")^2

# R-squared straight from the model
summary(fit.1)$r.squared
```

. . .

And the *t* for the slope?

```{r}
tidy(fit.1)
cor.test(cah$age, cah$transformers)$statistic
```

------------------------------------------------------------------------

## Significance $\neq$ importance

The $p$-value tells us the effect is **likely not zero**. It says *nothing* about whether the effect is large or meaningful.

::: callout-important
With a large enough sample, a tiny, trivial effect becomes statistically significant. Our *n* here is **938**.
:::

**Statistical Significance** $\neq$ **Practical Significance**

------------------------------------------------------------------------

## Effect size for a regression

1.  **The slope itself** — in the units of your study. The most concrete thing you have.

2.  **Standardized slope** — in SD units. Cohen's rough guide: small (\|.10\|), medium (\|.30\|), large (\|.50\|)

3.  **Coefficient of determination** $R^2$ — proportion of variance accounted for

. . .

For our model: $r = -.27$, so $R^2 = .071$

> "Approximately **7%** of the variability in Transformers movies seen is accounted for by age."

------------------------------------------------------------------------

## Writing it up

Template: *b* = slope, *t*(df) = statistic, *p* = p-value, plus $R^2$

> Age significantly predicted the number of Transformers movies a respondent had seen (*b* = −0.03, *p* \< .001). Each additional decade of age was associated with about a quarter of a movie fewer, and age accounted for 7.1% of the variance in movies seen ($R^2$ = .07).

. . .

::: callout-tip
Notice what the good version includes: the estimate **in real units**, the test, *and* a statement of how much it matters. All three.

Some folks will include the t-values, but I find that overwhelming. Just have that in the tables.
:::

------------------------------------------------------------------------

## Pretty tables with `sjPlot`

```{r}
tab_model(fit.1, show.std = TRUE)
```

------------------------------------------------------------------------

## Data, fitted, and residuals

```{r}
model_info <- augment(fit.1)
head(model_info)
```

::: notes
Point out that the average of the residuals is 0 — just like the average deviation from the mean is 0.
:::

------------------------------------------------------------------------

### The relationship between $\hat{Y_i}$ and $e_i$

```{r, fig.width=8, fig.height = 5}
#| code-fold: true

model_info %>%
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = .3) +
  geom_smooth(se = F, method = "lm") +
  scale_y_continuous("e") +
  scale_x_continuous(expression(hat(Y))) +
  theme_bw(base_size = 25)
```

. . .

Flat. Always. **By construction** — and in Week 10 we will use that fact to catch models that are lying to us.

# Part 4: The Lab 5 Punchline

------------------------------------------------------------------------

## A t-test *is* a regression

```{r}
cah_mf <- cah %>% filter(gender == "Male" | 
                           gender == "Female") 
#OR

cah_mf <- cah %>% filter(gender %in% c("Male", "Female"))


t.test(transformers ~ gender, data = cah_mf, var.equal = TRUE)
```

------------------------------------------------------------------------

## The same question as a line

```{r}
lm(transformers ~ gender, data = cah_mf) %>% tidy()
```

::: incremental
- The **intercept** (1.21) is the mean of the *first* group (Female)
- The **slope** (0.23) is the *difference* between the means
- Same *t*. Same *p*. One machine.
:::

------------------------------------------------------------------------

## Why this matters

::: callout-note
You do not have to memorize a separate procedure for every design.

`outcome ~ predictor` handles the continuous case **and** the group case. Over the next few weeks we just add more predictors to the same machine.
:::

**Next week:** groups as regression, in full — t-tests and ANOVA as `lm()`.

------------------------------------------------------------------------

## One more thing: regression to the mean

An observation about heights motivated the development of regression: if you selected an exceptionally tall (or short) parent, their child was almost always **less** extreme.

```{r,  warning = F, message = F, fig.height = 4}
#| code-fold: true

library(psychTools)
heights <- psychTools::galton
mod <- lm(child ~ parent, data = heights)
heights <- broom::augment(mod)

heights %>%
  ggplot(aes(x = parent, y = child)) +
  geom_jitter(alpha = .3) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_hline(aes(yintercept = 72), color = "red") +
  geom_vline(aes(xintercept = 72), color = "red") +
  theme_bw(base_size = 20)
```

------------------------------------------------------------------------

## Regression to the mean

**Regression to the mean**: a random variable that produces an extreme score on a first measurement tends to produce a less extreme score on a second measurement.

::: callout-warning
This is a threat to internal validity whenever an intervention is assigned **based on** first-measurement scores.

Tutoring the students who bombed the midterm? They would have improved somewhat *anyway*.
:::

------------------------------------------------------------------------

## 🛠️ Activity Stop 2: Parts B & C

::: callout-tip
**One Line, Two Stories** · \~35 min · Pairs
:::

**Part B (\~25 min):** Do older people read more books?

- You will get *p* = .027 — and then one person who reported reading **1,000 books** will make it disappear
- The p-value flips on a judgment call **you** make

**Part C (\~10 min):** The t-test / regression equivalence, in your own file

[Open the activity](../class-activities/wk5_linear-model.html){.button target="_blank"}

------------------------------------------------------------------------

## ✍️ Wrap-Up

Finish these at the bottom of your activity file:

1.  A slope, stated in the units of a study, **without using the word "slope."**
2.  One intercept from today that is meaningless, and why.
3.  One thing that is still muddy → put it in your journal this week.

------------------------------------------------------------------------

## For next time

📋 **Lab 6 — Reading the Line**

- Almost no coding. I give you the output; you tell me what it means.
- Uses the Lab 5 data plus a new context

📖 Chapter 15 — ST

::: callout-important
## Schedule change

The lab schedule has moved up from the PDF syllabus. Website = current version.
:::
