---
title: "Week 07: Variability & Model Fit"
subtitle: "Date: October 8, 2026"
footer:  "[course-website](https://dharaden.github.io/psyc640/)"
logo: "images/640_hex.png"
format:
  revealjs:
    theme: [default, brand, clean.scss]
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false
    chalkboard: true
    overview: false
    scrollable: true
    code-line-numbers: true
editor: visual
execute:
  echo: true
  freeze: auto
---

## Today...

**How good is this model, actually?**

- Where the variance goes: $SS_{total} = SS_{model} + SS_{residual}$
- R², one more time — and what it is *not*
- The residual standard error: your typical miss, in real units
- The F test as a ratio
- Why R² can only go **up**, and what to use instead
- 👀 A sneak peek at two predictors

```{r, results = 'hide', message = F, warning = F}
library(here)
library(tidyverse)
library(rio)
library(broom)
library(psych)

options(scipen = 999)

asap <- import(here("files", "data", "asap_data.csv"))
```

------------------------------------------------------------------------

## Our data: the ASAP study 😴

400 adolescents, measured on sleep, stress, and mood.

```{r}
asap %>%
  select(T1_SleepDur, T1_PeerStress, T1_AcadStress, T1_Dep) %>%
  describe(fast = TRUE)
```

> **Research question:** Does how long adolescents sleep predict depressive symptoms?

# Part 1: Where Does the Variance Go?

------------------------------------------------------------------------

## Start with the dumbest possible model

Suppose you had **no predictor at all**. What is your best guess for anyone's depression score?

. . .

The **mean**. That's it. That's the model.

```{r}
mean(asap$T1_Dep)
```

. . .

How wrong is that model? Add up every squared miss:

$$SS_{total} = \sum(Y_i - \bar{Y})^2 = 1840.5$$

This is the **total variability** we are trying to account for. Everything from here is about shrinking it.

------------------------------------------------------------------------

## Now add a predictor

```{r}
fit_sleep <- lm(T1_Dep ~ T1_SleepDur, data = asap)
tidy(fit_sleep)
```

. . .

Each additional hour of sleep is associated with **0.55 fewer** points of depression.

But our question today is different: **how much better is this line than the mean?**

------------------------------------------------------------------------

## Three distances, one observation

For any single adolescent:

::::: columns
::: {.column width="50%"}
$$\underbrace{Y_i - \bar{Y}}_{\text{total}} = \underbrace{\hat{Y_i} - \bar{Y}}_{\text{model}} + \underbrace{Y_i - \hat{Y_i}}_{\text{residual}}$$
:::

::: {.column width="50%"}
- **Total**: how far this person is from the overall mean
- **Model**: how far the *line* moved them from the mean
- **Residual**: how far the line still missed
:::
:::::

. . .

Square them, add them up across all 400 people, and the same decomposition holds:

$$SS_{total} = SS_{model} + SS_{residual}$$

------------------------------------------------------------------------

## See it in the numbers

```{r}
anova(fit_sleep)
```

::: incremental
- $SS_{model}$ = **121.2** — variability the line accounts for
- $SS_{residual}$ = **1719.4** — variability it does not
- Together: 121.2 + 1719.4 = **1840.5** = $SS_{total}$ ✅
:::

------------------------------------------------------------------------

## R² is just a share

$$R^2 = \frac{SS_{model}}{SS_{total}} = \frac{121.2}{1840.5} = .066$$

```{r}
glance(fit_sleep) %>% select(r.squared, adj.r.squared, sigma, statistic, p.value)
```

. . .

**6.6%** of the variability in depression is accounted for by sleep duration.

Which means **93.4% is not**. Both halves of that sentence are the finding.

------------------------------------------------------------------------

## Small R², strong evidence

::::: columns
::: {.column width="50%"}
**R² = .066**

Sleep duration tells us very little about any individual adolescent's depression.
:::

::: {.column width="50%"}
**p \< .001**

With n = 400, we are confident the slope is not zero.
:::
:::::

. . .

::: callout-important
These are answers to **different questions**. "Is there an effect?" and "how much does it explain?" are not the same, and a p-value only answers the first.
:::

. . .

This is normal in psychology. Behavior has many causes; one predictor rarely owns much of the variance.

# Part 2: Fit in Real Units

------------------------------------------------------------------------

## R² is unitless. That's a problem.

".066" does not tell you whether this model is useful for a clinician.

. . .

The **residual standard error** does. It is roughly the size of a typical miss, in the units of the outcome:

$$RSE = \sqrt{\frac{SS_{residual}}{n - k - 1}} = \sqrt{\frac{1719.4}{398}} = 2.08$$

```{r}
glance(fit_sleep)$sigma
sd(asap$T1_Dep)
```

------------------------------------------------------------------------

## The comparison that matters

::: incremental
- Guess the **mean** for everyone → you are off by about **2.15** points on average
- Use the **model** → you are off by about **2.08** points
:::

. . .

We bought ourselves about **0.07 points** of accuracy.

. . .

::: callout-tip
That is the honest way to describe a small R² to a non-statistician: *not* "6.6% of variance," which means nothing to most people, but "knowing how long a teenager slept barely improves our guess about their mood."
:::

------------------------------------------------------------------------

## The F test is a ratio

$$F = \frac{MS_{model}}{MS_{residual}} = \frac{\text{variance the model explains, per predictor}}{\text{variance it doesn't, per leftover df}}$$

```{r}
glance(fit_sleep) %>% select(statistic, df, df.residual, p.value)
```

::: incremental
- $F$ = 28.05 on 1 and 398 df
- If the model explained **nothing**, F would hover around 1
- With one predictor, $F = t^2$: $(-5.30)^2 = 28.05$ ✅ — the same test again
:::

# Part 3: Adding Predictors 👀

------------------------------------------------------------------------

## Does peer stress help?

```{r}
fit_two <- lm(T1_Dep ~ T1_SleepDur + T1_PeerStress, data = asap)
tidy(fit_two)
```

. . .

Both predictors are significant. The sleep slope barely moved (−0.55 → −0.50).

------------------------------------------------------------------------

## What happened to fit?

| Model               | R²       | Adjusted R² | RSE       |
|---------------------|----------|-------------|-----------|
| Sleep only          | .066     | .064        | 2.078     |
| Sleep + peer stress | **.093** | **.088**    | **2.051** |

::: incremental
- $\Delta R^2$ = **.027** — peer stress accounts for 2.7% *beyond* sleep
- The typical miss dropped from 2.078 to 2.051
- Both numbers point the same way: a modest improvement
:::

------------------------------------------------------------------------

## Now let's add something useless

```{r}
fit_three <- lm(T1_Dep ~ T1_SleepDur + T1_PeerStress + T1_AcadStress, data = asap)
tidy(fit_three)
```

Academic stress: *b* = −0.007, *p* = .90. As close to nothing as you will ever see.

------------------------------------------------------------------------

## Watch what R² does

| Model              | R²        | Adjusted R²  | RSE          |
|--------------------|-----------|--------------|--------------|
| Sleep only         | .0658     | .0635        | 2.078        |
| \+ peer stress     | .0928     | .0882        | 2.051        |
| \+ academic stress | **.0928** | **.0859** ⬇️ | **2.053** ⬆️ |

::: incremental
- R² **did not go down** — it never can
- Adjusted R² **did**, and the typical miss got **worse**
:::

------------------------------------------------------------------------

## Why R² can only climb

Adding a predictor gives OLS one more knob to turn. At worst it sets the coefficient near zero and fit stays the same; usually it catches a sliver of noise and R² ticks up.

. . .

$$R^2_{adj} = 1 - (1 - R^2)\frac{n - 1}{n - k - 1}$$

Every predictor $k$ costs you something. Adjusted R² only rises if the new variable earns its place.

::: callout-important
**R² is not a model-selection tool.** The biggest R² always belongs to the model with the most predictors. Week 11 gives us honest ways to compare models.
:::

------------------------------------------------------------------------

## ⏱️ One last thing: look at the residuals {background-color="#EFE6F7"}

We spent all day computing $SS_{residual}$. Let's actually **look** at the residuals instead of just squaring them.

```{r, fig.height=3.4}
#| code-fold: true
augment(fit_two) %>% 
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = .4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(se = FALSE, color = "#B3261E") +
  labs(x = "Fitted (predicted depression)", y = "Residual") +
  theme_bw(base_size = 16)
```

A shapeless cloud centered on zero. That is what "nothing obviously wrong" looks like.

------------------------------------------------------------------------

## What you are glancing for

```{r, fig.height=4.2}
#| code-fold: true
set.seed(640)
n <- 200
x <- runif(n, 0, 10)

sims <- bind_rows(
  tibble(kind = "1. Healthy",          x = x, y = 2 + 0.8 * x + rnorm(n, 0, 2)),
  tibble(kind = "2. Curve (wrong shape)", x = x, y = 2 + 0.8 * x^2 + rnorm(n, 0, 6)),
  tibble(kind = "3. Funnel (spread grows)", x = x, y = 2 + 0.8 * x + rnorm(n, 0, 0.4 * x)),
  tibble(kind = "4. One strange point",  x = x, y = 2 + 0.8 * x + rnorm(n, 0, 2))
) %>% 
  mutate(y = if_else(kind == "4. One strange point" & row_number() == n(), y + 40, y))

sims %>% 
  group_by(kind) %>% 
  group_modify(~ augment(lm(y ~ x, data = .x))) %>% 
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = .4, size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ kind, scales = "free") +
  labs(x = "Fitted", y = "Residual") +
  theme_bw(base_size = 13)
```

::: incremental
- **Curve** → the line is the wrong shape for these data
- **Funnel** → the model is more wrong for some people than others
- **Strange point** → one observation may be running the whole model (remember the 1,000-books reader?)
:::

. . .

::: callout-note
That is the entire diagnostic toolkit for now: **plot the residuals, look for a pattern.** Week 10 turns each of these into a named assumption with a proper test. For the midterm, the glance is enough.
:::

------------------------------------------------------------------------

## 🛠️ Activity: Does the Algorithm Know What a Hit Is?

::: callout-tip
**Week 7 Exercises** · \~45 min · Pairs
:::

Spotify data: 952 popular songs, with danceability, energy, valence, acousticness — and streams.

- Four audio features together explain about **1%** of streams
- One boring variable explains **62%**
- You will find it, and then explain why the fancy features lost

[Open the activity](../class-activities/wk7_fit.html){.button target="_blank"}

------------------------------------------------------------------------

## ✍️ Wrap-Up

Finish these at the bottom of your activity file:

1.  In one sentence, explain R² to someone who has never taken statistics.
2.  Why can a predictor be statistically significant and still barely improve a model?
3.  One thing that is still muddy → journal.

------------------------------------------------------------------------

## For next time

📋 **Lab 8 — How Good Is This Model?**, due **Tuesday 10/13 at 11:59pm**

::: callout-warning
**10/13 is during Fall Break.** The lab is submitted online as usual — nothing to hand in physically, and no class that Monday or Tuesday.
:::

- Model fit on a real anxiety dataset, in the units people care about
- Ends with a **sneak peek**: one predictor that looks useless alone, and what happens when you add a second

📖 Chapter 10 — ROS (for next week's multiple regression)

::: callout-important
## 📂 Midterm Project — posted today

The [**Midterm Project**](../labs/midterm-project.html) is up on the website now. **Due Tuesday, 10/27 at 11:59pm.** It is worth 200 points (20% of your grade).

- **Parts 1-5 use only what you already know** - cleaning, scoring, describing, one-predictor models. You can start this weekend.
- **Part 6 needs multiple regression**, which is next class (10/15).
- You pick your own research question. Read the whole thing before you leave today.

No lab is due in Weeks 9 or 10, so that time is yours for the project.
:::
