---
title: "Week XX: TITLE"
subtitle: "Date: MM-DD"
footer:  "[course-website](https://dharaden.github.io/psyc640/)"
logo: "images/640_hex.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
    overview: true
    code-line-numbers: true
editor: visual
execute:
  freeze: auto
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)

options(scipen = 100)
```

# Today

- Goals
- Why it matters (results-section lens)
- What you’ll be able to do by the end


# Warm-up

- 2-minute think-pair-share: interpret this figure/table snippet.


# Concepts (brief)

- Keep math minimal; emphasize assumptions, use-cases, effect sizes.


# Live demo (R)

```{r}
# setup
library(rio)
library(here)
#dat <- import(here::here("data","sample.csv"))
```


# Guided practice

- Students follow along: 3–5 tasks with immediate feedback.


# Interpretation

- Translate model output to 2–3 sentences as in a results section.


# Exit ticket

- One key takeaway, one question.


# Next

- Prep and due items for next week.
