---
title: "Week 06: Comparing Groups as Regression"
subtitle: "Date: October 1, 2026"
footer:  "[course-website](https://dharaden.github.io/psyc640/)"
logo: "images/640_hex.png"
format:
  revealjs:
    theme: [default, brand, clean.scss]
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false
    chalkboard: true
    overview: false
    scrollable: true
    code-line-numbers: true
editor: visual
execute:
  echo: true
  freeze: auto
---

## Today...

**Groups, as lines**

- Why a t-test *is* a regression (we prove it properly this time)
- Dummy coding: how R turns a group into a number
- Three or more groups: ANOVA is the same machine
- Why not just run a bunch of t-tests?
- Effect size: *d*, R², and which one to report

```{r, results = 'hide', message = F, warning = F}
library(here)
library(tidyverse)
library(rio)
library(broom)
library(psych)
library(emmeans)

options(scipen = 999)

tipi <- import(here("files", "data", "final_tipi.csv")) %>%
  # case_when(variable of interest == old_value ~ new value)
  mutate(year_group = case_when(year_school == 1 ~ "First year",
                                year_school == 2 ~ "Sophomore",
                                year_school >= 3 ~ "Junior+"))
```

------------------------------------------------------------------------

## Where we left off

Last week, in Part C of the activity:

::: incremental
- You ran a **t-test** on two groups and got *t* = −2.29, *p* = .022
- You ran `lm()` on the same two groups and got **the same** *t* and *p*
- The **intercept** turned out to be one group's mean, and the **slope** the difference between the means
:::

. . .

Today: *why*, and what happens when there are **three** groups.

------------------------------------------------------------------------

## Our data:  😴🏫

We are going back to the **TIPI dataset** from Week 3 — the one with 99 students, their personality scores, and how much they sleep.

```{r}
tipi %>%
  count(year_group)
```

> **Research question:** Do students further along in school sleep differently than first-years?

# Part 1: Two Groups

------------------------------------------------------------------------

## Start with two

```{r}
two_years <- tipi %>%
  filter(year_group %in% c("First year", "Junior+"))

two_years %>%
  group_by(year_group) %>%
  summarize(n = n(),
            M = mean(hours_of_sleep),
            SD = sd(hours_of_sleep))
```

. . .

Juniors and above sleep about **0.79 hours** — roughly 47 minutes — more per night. Is that more than noise?

------------------------------------------------------------------------

## The test you already know

```{r}
t.test(hours_of_sleep ~ year_group, data = two_years, var.equal = TRUE)
```

*t*(74) = −3.38, *p* = .001.

Students ranked as juniors or higher reported significantly greater amounts of sleep (M = 6.92, SD = 0.64) compared to freshman (M = 6.13, SD = 1.28; *p* = .001).

------------------------------------------------------------------------

## The same question as a line

```{r}
mod_two <- lm(hours_of_sleep ~ year_group, data = two_years)
summary(mod_two)
```

. . .

::: callout-important
Same *t* (the sign just depends on which group R treats as "first"). Same *p*. The t-test is a regression with **one categorical predictor**.
:::

------------------------------------------------------------------------

## But how do you put a *group* in an equation?

$$\hat{Y_i} = b_0 + b_1X_i$$

$X$ has to be a **number**. "Junior+" is not a number.

. . .

So we make one. **Dummy coding**:

| Group      | coded as |
|------------|----------|
| First year | **0**    |
| Junior+    | **1**    |

That's it. That is the whole trick.

------------------------------------------------------------------------

## Read the equation twice

$$\widehat{sleep} = 6.13 + 0.79(junior^+)$$

::::: columns
::: {.column width="50%"}
**First year** → plug in 0

$$6.13 + 0.79(0) = 6.13$$

The **intercept** is the mean of the group coded 0.
:::

::: {.column width="50%"}
**Junior+** → plug in 1

$$6.13 + 0.79(1) = 6.92$$

The **slope** is the **difference** between the group means.
:::
:::::

. . .

A "line" through a predictor with two values can only ever produce **two** predicted values: the two group means.

------------------------------------------------------------------------

## The group that gets the 0 is the *reference*

R picks it for you: **alphabetically**. "First year" beats "Junior+".

```{r}
two_years %>%
  mutate(year_group = fct_relevel(year_group, "Junior+")) %>%
  lm(hours_of_sleep ~ year_group, data = .) %>%
  summary()
```

. . .

The intercept becomes 6.92, the slope flips to −0.79, and *t*, *p*, R² and every residual stay exactly where they were.

------------------------------------------------------------------------

## What about Welch?

`t.test()` defaults to **Welch's** t-test (unequal variances). `lm()` always assumes **equal** variances, like Student's t-test.

. . .

Look back at those SDs: **1.28** for first-years, **0.64** for juniors+. The first-years are twice as spread out.

::: callout-note
That is exactly the situation Welch was built for. Here the groups are nearly the same size (39 vs 37), so the two tests agree — but with unequal *n* **and** unequal spread, they would not. Know which one you ran.
:::

------------------------------------------------------------------------

## Effect size: two flavors

```{r}
effectsize::cohens_d(hours_of_sleep ~ year_group, data = two_years)
glance(mod_two) %>% select(r.squared, statistic, p.value)
```

::: incremental
- **Cohen's *d* = −0.78** (0.78 in magnitude; the sign just reflects group order) — the difference in standard deviations. "How separated are the groups?"
- **R² = .133** — the proportion of variance in sleep accounted for by year group. "How much does knowing the group help?"
:::

. . .

Same model. Report *d* for a two-group comparison; R² scales up to everything we do next.

------------------------------------------------------------------------

## 🛠️ Activity Stop 1: Same Question, Three Ways

::: callout-tip
**Part A: Follow along** · \~20 min · Pairs
:::

New data — the Pokémon set from your in-class check:

- Run the t-test, then the regression, on the same two groups
- Find the intercept in the group means
- Flip the reference group and predict what will change **before** you run it

[Open the activity](../class-activities/wk6_groups.html){.button target="_blank"}

------------------------------------------------------------------------

## ☕ Break {background-color="#C9ADE0"}

**10 minutes**

When we come back: **three** groups.

# Part 2: Three or More Groups

------------------------------------------------------------------------

## Put the sophomores back

```{r}
tipi %>%
  group_by(year_group) %>%
  summarize(n = n(),
            M = mean(hours_of_sleep),
            SD = sd(hours_of_sleep))
```

. . .

Sophomores land in the middle. Now we have **three** means — and three possible comparisons.

------------------------------------------------------------------------

## Why not just run three t-tests? 🤔

first-year vs sophomore · first-year vs junior+ · sophomore vs junior+

. . .

Each test has a 5% false-positive rate. We set that when we accept our cutoff of p \< .05. If the three tests were **independent**, the chance of at least one false positive would be:

$$1 - (1 - .05)^3 = .143$$

. . .

They are *not* independent — the three tests reuse the same three groups — so that formula overstates it a bit. Let's just **check**.

------------------------------------------------------------------------

## Build a world where nothing is true

Same idea as the Inference Machine in Week 4: simulate studies where we **know** the null is true, then count how often we fool ourselves.

```{r}
#| code-fold: true
set.seed(640)

one_study <- function(n = 30) {
  # three groups drawn from the SAME population: no real differences
  a <- rnorm(n)
  b <- rnorm(n)
  c <- rnorm(n)
  
  # run all three pairwise t-tests
  p_ab <- t.test(a, b, var.equal = TRUE)$p.value
  p_ac <- t.test(a, c, var.equal = TRUE)$p.value
  p_bc <- t.test(b, c, var.equal = TRUE)$p.value
  
  # keep the smallest p-value
  min(p_ab, p_ac, p_bc)
}

smallest_p <- replicate(1000, one_study())

# how often do we get at least one "significant" result by chance?
mean(smallest_p < 0.05)
```

------------------------------------------------------------------------

## 1,000 studies where nothing was going on

```{r, fig.height=3.6}
#| code-fold: true
tibble(smallest_p = smallest_p) %>% 
  ggplot(aes(x = smallest_p, fill = smallest_p < .05)) +
  geom_histogram(binwidth = .025, boundary = 0, color = "white") +
  geom_vline(xintercept = .05, linetype = "dashed") +
  scale_fill_manual(values = c("grey70", "#B3261E"), guide = "none") +
  labs(x = "Smallest p-value in the study", y = "Number of studies",
       title = "Three groups, three t-tests, zero real differences") +
  theme_bw(base_size = 16)
```

About **12%** of these studies produced at least one "significant" difference. None of them were real.

------------------------------------------------------------------------

## It gets worse fast

```{r}
#| code-fold: true
set.seed(640)

num_groups <- 5
sample <- 30

one_study <- function(n = 30, k = 3) {
  # k groups, all drawn from the SAME population - no real differences
  d <- tibble(group = rep(LETTERS[1:k], each = n),
              y     = rnorm(n * k))
  
  # run every pairwise t-test, keep the smallest p-value
  combn(LETTERS[1:k], 2, simplify = FALSE) %>% 
    map_dbl(function(pair) {
      t.test(y ~ group, data = filter(d, group %in% pair), 
             var.equal = TRUE)$p.value
    }) %>% 
    min()
}

smallest_p <- map_dbl(1:1000, function(i) one_study(sample, num_groups))

prop_sig <- mean(smallest_p < .05)

tibble(smallest_p = smallest_p) %>% 
  ggplot(aes(x = smallest_p, fill = smallest_p < .05)) +
  geom_histogram(binwidth = .025, boundary = 0, color = "white") +
  geom_vline(xintercept = .05, linetype = "dashed") +
  scale_fill_manual(values = c("grey70", "#B3261E"), guide = "none") +
  labs(x = "Smallest p-value in the study", 
       y = "Number of studies",
       title = paste(num_groups, "groups,", choose(num_groups, 2),
       "t-tests, zero real differences"), 
       subtitle = paste0(round(prop_sig * 100, 1),
                                    "% of studies found p < .05")) +
  theme_bw(base_size = 16)
```

------------------------------------------------------------------------

## It gets worse fast

| Groups | Pairwise tests | False-positive rate      |
|--------|----------------|--------------------------|
| 2      | 1              | **4.9%** ← as advertised |
| 3      | 3              | **12.3%**                |
| 4      | 6              | **20.2%**                |
| 5      | 10             | **29.2%**                |

[*20,000 simulated studies per row, n = 30 per group, all groups from the same population.*]{style="font-size:0.7em"}

::: incremental
- The naive formula ($1-.95^{10}$ = 40% for five groups) **overstates** it, because the tests overlap
- But the honest number is still **six times** the rate you think you are running
:::

. . .

This is why we start with **one** omnibus test, and why pairwise comparisons get **corrected**.

------------------------------------------------------------------------

## Two groups needed one dummy. Three need two.

| Group                      | `Junior+` | `Sophomore` |
|----------------------------|-----------|-------------|
| **First year** (reference) | 0         | 0           |
| Junior+                    | 1         | 0           |
| Sophomore                  | 0         | 1           |

::: incremental
- $k$ groups → $k - 1$ dummy variables
- Every group gets a unique pattern of 0s and 1s
- The reference group is the one that is **all zeros**
:::

. . .

R builds these columns for you the moment it sees a factor.

------------------------------------------------------------------------

## The model

```{r}
mod_year <- lm(hours_of_sleep ~ year_group, data = tipi)
summary(mod_year)
```

------------------------------------------------------------------------

## Reading it

$$\widehat{sleep} = 6.13 + 0.79(junior^+) + 0.35(sophomore)$$

::: incremental
- **Intercept 6.13** = mean sleep of **first-years** (the reference)
- **+0.79** = juniors+ sleep 0.79 hours more **than first-years**, *t*(96) = 3.47, *p* \< .001
- **+0.35** = sophomores sleep 0.35 hours more **than first-years**, *t*(96) = 1.34, *p* = .184
- Every coefficient is a comparison **to the reference group** — not to each other
:::

. . .

So where is sophomore vs junior+? **It isn't in this table.**

------------------------------------------------------------------------

## The omnibus F test

At the bottom of that output:

> F(2, 96) = 6.02, **p = .003**, R² = **.111**

::: incremental
- $H_0$: all three population means are equal
- A small *p* says **at least one** group differs — it does not say which
- This is the ANOVA you already know. Same model, different summary.
:::

```{r}
anova(mod_year)
```

------------------------------------------------------------------------

## ANOVA *is* regression

```{r}
# the classic way
summary(aov(hours_of_sleep ~ year_group, data = tipi))
```

. . .

Same F. Same p. `aov()` is a wrapper around `lm()` that prints a sums-of-squares table instead of coefficients.

::: callout-note
R² here is what ANOVA calls $\eta^2$ (eta squared): **11.1%** of the variance in sleep is accounted for by year in school.
:::

------------------------------------------------------------------------

## Getting the comparison the table hid

```{r}
emmeans(mod_year, pairwise ~ year_group, adjust = "none")$contrasts
```

::: incremental
- First year vs Junior+: **significant** (Tukey *p* = .002)
- First year vs Sophomore: not significant (*p* = .38)
- Sophomore vs Junior+: not significant (*p* = .22)
:::

. . .

The omnibus test was significant because of **one** of the three comparisons. Without the pairwise step you would not know which.

------------------------------------------------------------------------

## Writing it up

> Average nightly sleep differed significantly across year in school (*F*(2, 96) = 6.02, *p* = .003, $\eta^2$ = .11). First-year students reported the least sleep (*M* = 6.13, *SD* = 1.28), followed by sophomores (*M* = 6.48, *SD* = 0.90) and juniors or above (*M* = 6.92, *SD* = 0.64). Tukey-adjusted comparisons indicated that juniors and above slept significantly more than first-year students (*p* = .002); no other pairwise comparison was significant.

. . .

::: callout-tip
Notice the shape: **omnibus test → group descriptives → which pairs differ**. In that order, every time.
:::

------------------------------------------------------------------------

## And the obvious caveat

Nobody was **randomly assigned** to be a first-year.

::: incremental
- Maybe upper-year students have more control over their schedules
- Maybe the heavy-sleep-deprivation majors drop out before junior year
- Maybe first-years live in noisier housing
:::

. . .

The model is fine. The **causal claim** is not available — same lesson as Lab 6. Next week's lab is the exception that proves the rule: a study where people *were* randomly assigned.

------------------------------------------------------------------------

## Did we forget to visualize??

```{r}
tipi %>% 
  ggplot(aes(year_group, hours_of_sleep, 
             fill = year_group)) + 
  geom_violin() + 
  geom_jitter(width = .1) + 
  labs(title = "Hours of Sleep by Year in School", 
       x = "Year", 
       y = "Hours of Sleep") + 
  theme(legend.position = "none")
```

------------------------------------------------------------------------

## The big picture so far

| What you want        | The model     | What people call it              |
|----------------------|---------------|----------------------------------|
| Continuous predictor | `y ~ x`       | Regression                       |
| 2 groups             | `y ~ group`   | t-test                           |
| 3+ groups            | `y ~ group`   | One-way ANOVA                    |
| Several predictors   | `y ~ x1 + x2` | **Multiple regression (Week 8)** |

. . .

One machine. You are just changing what goes on the right-hand side.

------------------------------------------------------------------------

## 🛠️ Activity Stop 2: Parts B & C

::: callout-tip
**Same Question, Three Ways** · \~30 min · Pairs
:::

**Part B:** three Pokémon types, from descriptives to a written-up result

**Part C:** choose a reference group on purpose, and see what changes (and what doesn't)

[Open the activity](../class-activities/wk6_groups.html){.button target="_blank"}

------------------------------------------------------------------------

## For next time

📋 **Lab 7 — Does the Therapy Work?**, due **Tuesday 10/6 at 11:59pm**

- A real three-arm randomized trial: Individual CBT, Group CBT, Waitlist
- You fit the model, pick the reference group, and write the results
- Five participants dropped out — you will have to decide what to say about that

📖 Chapter 8 — ROS · Chapters 5 & 14.5 — ST · Chapter 15.4 — LSR

::: callout-note
Heads up: the **midterm project** is handed out in two weeks (10/15) and due 10/27. Lab 8 is due Tue 10/13, which is during Fall Break — submit it online as usual.
:::
