---
title: "Week 01: Getting Started!"
subtitle: "Date: 08-25"
footer:  "[course-website](https://dharaden.github.io/psyc640/)"
logo: "images/640_hex.png"
format: 
  revealjs:
    theme: [default, seasons.scss]
    multiplex: false
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
    overview: true
    code-line-numbers: true
editor: visual
execute:
  freeze: auto
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)

options(scipen = 100)
```

# Today {.fall}

-   Goals
-   Why it matters (results-section lens)
-   What you’ll be able to do by the end

# Warm-up {.fall}

-   2-minute think-pair-share: interpret this figure/table snippet.

# Concepts (brief) {.fall}

-   Keep math minimal; emphasize assumptions, use-cases, effect sizes.

# Live demo (R) {.fall}

```{r}
# setup
library(rio)
library(here)
#dat <- import(here::here("data","sample.csv"))
```

# Guided practice {.fall}

-   Students follow along: 3–5 tasks with immediate feedback.

# Interpretation {.fall}

-   Translate model output to 2–3 sentences as in a results section.

# Exit ticket {.fall}

-   One key takeaway, one question.

# Next {.fall}

-   Prep and due items for next week.
