---
title: "Reviews & Archival Data"
subtitle: "PSYC 641 - Spring 2025"
footer:  "[course-website](https://dharaden.github.io/psyc641-s25/)"
logo: "images/641_hex.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
editor: visual
execute:
  freeze: auto
---

```{r setup}
#| include: false

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)

options(scipen = 100)
```

```{r packages}
#| echo: false
#| message: false

# load packages
library(countdown)  
```

## Reminders/Updates

::: nonincremental
-   Data Analytic Plan - Due April 13

    -   [Check it out](/project/final-proj-desc.html)

-   Final Reports and Presentations - Due on April 28

    -   Updating and combining all that we've done!
:::

------------------------------------------------------------------------

## Overview for Today

::: nonincremental
-   Review Articles

    -   Systematic & Scoping Reviews

    -   Meta-Analysis

-   Archival Data

    -   Where is it?

    -   Integrated Data Analysis?
:::

------------------------------------------------------------------------

## Review of the Literature

A formalized process of examining a body of research and synthesizing the findings.

Can be organized in a number of different ways:

::: nonincremental
-   Scoping Review

-   Systematic Review

-   Meta-Analysis
:::

------------------------------------------------------------------------

### Class Activity

Get onto groups of 2-3

Identify a Review or Meta-analysis in one of the following areas. Modify the research question as needed.

Fill out this [Google Sheet](https://docs.google.com/spreadsheets/d/1z9ic3cYilVNrLHqUqzUzJ1pHkzsjbKegcIR2Sigwt1g/edit?usp=sharing){target="_blank"}

::: nonincremental
-   **Clinical** - The impact of a treatment on a form of psychopathology (e.g., IPT on Depression)
-   **Cognitive** - The role of executive function in task completion
-   **Social** - The role of information source on attitude change
-   **UX/IO** - The impact of light vs. dark mode on productivity
-   **Other** - You choose!
:::

```{r}
countdown(minutes = 10)
```

------------------------------------------------------------------------

### Types of Reviews

**Scoping Review**

::: nonincremental
-   A broad qualitative overview of an area that reflects on methodologies and procedures
:::

**Systematic Review**

::: nonincremental
-   More specific than scoring review. Focuses on a single question and qualitatively analyses research.
:::

**Meta-Analysis**

-   Extension of systematic review that examines the information *quantitatively*

------------------------------------------------------------------------

### Class Check-in

Look to the Procedures section and identify:

-   Search Terms used

-   Inclusion/Exclusion Criteria

-   The [PRISMA](https://www.prisma-statement.org/){target="_blank"} flow-chart

------------------------------------------------------------------------

[A Beginners Guide to Systematic Reviews and Meta-Analyses](https://noah-schroeder.github.io/reviewbook/index.html){target="_blank"}

[metafor package in R](https://www.metafor-project.org/doku.php/help){target="_blank"}

[Doing Meta-Analysis in R](https://bookdown.org/MathiasHarrer/Doing_Meta_Analysis_in_R/){target="_blank"}
