---
title: "Experimental Design"
subtitle: "PSYC 641 - Spring 2025"
footer:  "[course-website](https://dharaden.github.io/psyc641-s25/)"
logo: "images/641_hex.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
editor: visual
execute:
  freeze: auto
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)

options(scipen = 100)
```

# Welcome Back!

## Updates/Check-in {.smaller}

-   [Upcoming Schedule](https://dharaden.github.io/psyc641-s25/)

-   [Final Project](https://dharaden.github.io/psyc641-s25/project/final-proj-desc.html):

    ::: nonincremental
    -   Up Next - Methods/Procedures draft
        -   Let's Talk about it
    :::

## Computational setup (don't think we'll need)

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)   # for data wrangling and visualization
library(scales)      # for pretty axis labels
library(knitr)       # for pretty tables
library(patchwork)   # for laying out plots

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 20))
```

# Experimental Designs

Identified as the "Gold Standard" within research. Why?

The ability to draw *causal* connections between variables

-   e.g., Diffusion of Responsibility (Darley & Latane, 1968)

# Causation

How can we determine causation? i.e., one variable **causes** another to change

## Shadish, Cook & Campbell (2002)

Drawing from John Stuart Mill’s philosophy

Three Criteria for Causal Relationships:

1.  **Temporal precedence** (cause precedes effect)

2.  **Covariation** (cause and effect are correlated)

3.  **Elimination of plausible alternative explanations** (ruling out confounding variables)

### Counterfactual Reasoning

Intro to Experimental Designs

Random Assignment being undesirable

What is Causality (Shadish Cook & Campbell)

Randomized Control Trials

-   What are the expectations based on the groups that are being compared

    -   What is the control group (treatment as usual, placebo)

-   Placebo Effect

    -   Depression Symptoms

-   When 9 is Larger than 221 (structure of the study)

-   Mundane Realism vs. Psychological Realism

```{r}
countdown(minutes = 8)
```

## 
