---
title: "missing-data-sim"
format: html
---

<https://stefvanbuuren.name/fimd/sec-MCAR.html>

-   [Reading Question Check-in](https://docs.google.com/document/d/1o9wHlyPcx_WAHtq5BVaz4JeqYU7oaHCVLzlZLiXw_4Y/edit?usp=sharing)

## Load Required Packages

```{r}
library(tidyverse)
library(mice)
library(broom)
library(naniar) # For visualizing missingness
```

## **Missing Data Activity**

**Background**\
You are analyzing data from a study investigating how **sleep quality** and **stress** influence **academic performance (GPA)** in high school students.

Variables:

-    Sleep quality (actigraphy-based, continuous)

-    Stress level (self-report scale, continuous)

-   GPA (end-of-semester)

Three missing data scenarios were created:

1.   **MCAR**: Some sleep trackers were lost at random.

2.  **MAR**: Students with higher stress levels are less likely to wear the sleep tracker.

3.   **MNAR**: Students with low GPA are more likely to skip GPA reporting.

### **Part 1: Explore the Missingness**

-   Use `naniar::gg_miss_var()` to visualize which variables have missingness and how much.

-   Which variables are most affected in each scenario?

-   How would this affect your analysis if you used listwise deletion?

### **Part 2: Analyze Using Different Methods**

For each dataset (MCAR, MAR, MNAR):

-   Fit a linear model (`y ~ x1 + x2`) using **listwise deletion**. Save the coefficients.

-   Use **mean imputation** to replace missing values in `x1` and `x2`, then refit the model. Save results.

-   Use **multiple imputation** via `mice`, then pool and extract estimates. Save results.

-   Compare your coefficients to the full (complete data) model. Which method is closest to the true values?

### **Part 3: Reflect and Discuss**

Which method performed best for each scenario? Why?

Which scenario would you feel most confident analyzing and reporting? Least confident? Explain.

Ethical question:

MNAR Case: GPA missing due to low performance

Is it ethical to impute GPA?

-   Could it misrepresent actual student needs?

-   How would decisions based on imputed data affect interventions?

## Simulate Dataset

You are conducting a study examining how **sleep quality and stress levels** predict **academic performance** in a sample of high school students.

```{r}
set.seed(42)

#sample Size
n <- 500

sim_data <- tibble(
  sleep_qual = rnorm(n),
  stress = rnorm(n),
  acad_perf  = 0.5 * sleep_qual + 0.3 * stress + rnorm(n)
)

```

### Missing Data Scenarios

```{r}
# MCAR: Sleep Quality missing at random
data_mcar <- sim_data %>%
  mutate(sleep_qual = if_else(row_number() %in% sample(1:n, 100), NA_real_, sleep_qual))

# MAR: Stress missing based on Sleep Quality
data_mar <- sim_data %>%
  mutate(prob_miss = plogis(sleep_qual),
         stress = if_else(runif(n) < prob_miss, NA_real_, stress)) %>%
  select(-prob_miss)

# MNAR: Academic Performance missing based on Academic Performance itself
data_mnar <- sim_data %>%
  mutate(prob_miss = plogis(acad_perf),
         acad_perf = if_else(runif(n) < prob_miss, NA_real_, acad_perf)) %>%
  select(-prob_miss)

```

### Fit Models

```{r}
# Function to fit a model and extract coefficients
fit_model <- function(data, formula = acad_perf ~ sleep_qual + stress) {
  lm(formula, data = data) %>%
    tidy()
}

```

#### Full Data

```{r}
full_data <- sim_data %>% 
  fit_model()
```

#### Listwise Deletion

```{r}
listwise_mcar <- data_mcar %>% 
  drop_na() %>% 
  fit_model()

listwise_mar  <- data_mar  %>% 
  drop_na() %>% 
  fit_model()

listwise_mnar <- data_mnar %>% 
  drop_na() %>% 
  fit_model()


```

Remove rows with missing values.

Simple and easy\
Reduces sample size and can introduce bias

#### Mean Imputation

Only doing this for Sleep Quality and Stress variables

```{r}
# Function for mean imputation
mean_impute <- function(data) {
  data %>%
    mutate(sleep_qual = if_else(is.na(sleep_qual), 
                        mean(sleep_qual, na.rm = TRUE), 
                        sleep_qual),
           stress = if_else(is.na(stress), 
                            mean(stress, na.rm = TRUE), 
                            stress))
}

## Fitting the models
mean_mcar <- data_mcar %>% 
  mean_impute() %>% 
  fit_model()

mean_mar  <- data_mar  %>% 
  mean_impute() %>% 
  fit_model()

mean_mnar <- data_mnar %>% 
  mean_impute() %>% 
  fit_model()

```

Replace missing values with the mean.

Easy to implement\
Underestimates variability, can bias estimates

#### Multiple Imputation (mice)

```{r}
mice_fit <- function(data, m = 5) {
  imp <- mice(data, m = m, printFlag = FALSE)
  pool(with(imp, lm(acad_perf ~ sleep_qual + stress))) %>% tidy()
}

mi_mcar <- mice_fit(data_mcar)
mi_mar  <- mice_fit(data_mar)
mi_mnar <- mice_fit(data_mnar)

```

Use `mice` to create and pool multiple imputed datasets.

Preserves variability and uncertainty\
More accurate with MAR data\
Assumes correct imputation model

### Combine Results

```{r}
bind_rows(
  full_data %>% mutate(method = "none", scenario = "Full"),
  listwise_mcar %>% mutate(method = "Listwise", scenario = "MCAR"),
  listwise_mar  %>% mutate(method = "Listwise", scenario = "MAR"),
  listwise_mnar %>% mutate(method = "Listwise", scenario = "MNAR"),
  mean_mcar     %>% mutate(method = "Mean", scenario = "MCAR"),
  mean_mar      %>% mutate(method = "Mean", scenario = "MAR"),
  mean_mnar     %>% mutate(method = "Mean", scenario = "MNAR"),
  mi_mcar       %>% mutate(method = "MI", scenario = "MCAR"),
  mi_mar        %>% mutate(method = "MI", scenario = "MAR"),
  mi_mnar       %>% mutate(method = "MI", scenario = "MNAR")
) %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = scenario, y = estimate, fill = method)) +
  geom_col(position = "dodge") +
  facet_wrap(~term) +
  labs(title = "Comparison of Coefficient Estimates",
       y = "Estimate", x = "Missing Data Scenario") +
  theme_minimal()

```

## Realistic Missingness Scenarios

#### **MCAR (Missing Completely at Random)**

Some students accidentally lost their actigraphy devices, and their sleep data is missing. This happened randomly across the sample.

**Implication**: No systematic bias, but listwise deletion reduces power.

#### **MAR (Missing at Random)**

Students with **higher stress** were more likely to skip the sleep tracking part of the study because they felt overwhelmed. As a result, **sleep data is more likely to be missing for students with high stress**.

**Implication**: Bias is introduced unless the stress variable is used during imputation.

#### **MNAR (Missing Not at Random)**

Students who were **failing classes** were more likely to **avoid reporting GPA** due to embarrassment or fear of consequences. Thus, missingness in GPA is directly related to GPA itself.

**Implication**: Hardest case—there's no observed variable to "explain" the missingness, and imputations are highly assumption-dependent.
