---
title: " "
author: "Me"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{r}

library(easystats)
library(dplyr)
# File management
library(here)
# for dplyr, ggplot2
library(tidyverse)
#Loading data
library(rio)
# Estimating Marginal Means
library(emmeans)
# Pretty Tables
library(kableExtra)
# Pretty variable names
library(janitor)

#Remove Scientific Notation 
options(scipen=999)

#tell R not to display scientific notation
options(scipen=999)

# Create a list of citations of the libraries that are used
report::cite_packages()

# Identify the version of R and the type of machine that is running it
report::report_system()

# Create a list of the package names and the in-text citation 
report::report_packages()
```

## Ghost Data

Since there are 2 header rows, we need to only include the first one. To
do that, we have to "skip" the first two and then give the names of the
columns back to the data.

We also need to specify what the missing values are. Typically we have
been working with `NA` which is more traditional. However, missing
values in this dataset are `DK/REF` and a blank. This will need to be
specified in the import function ([used this
website](https://epirhandbook.com/new_pages/importing.html#manage-a-second-header-row){target="_blank"})

```{r}
# get the names of your columns which is the first row
ghost_data_names <- read_csv(here("lectures", "data", "ghosts.csv")) %>% 
  names()

# import second time; skip row 2, and assign column names to argument col_names =
ghost_data <- read_csv(here("lectures", "data", "ghosts.csv"),
                       skip = 2,
                       col_names = ghost_data_names,
                       na = c("DK/REF", "", " ")
                      ) %>% 
  clean_names()
```



### Running the Test

Let's take a look at Income by Political Affiliation

```{r}
aov1 <- aov(income ~ political_affiliation, 
            data = ghost_data)

summary(aov1)
```


### Post-hoc tests

```{r}

#get basic summary stats
ghost_data %>% 
  group_by(political_affiliation) %>% 
  summarise(mean = mean(income, na.rm = TRUE), 
            sd = sd(income, na.rm = TRUE))


# conduct the comparisons
emmeans(aov1, pairwise ~ political_affiliation, 
        adjust = "none")
```