---
title: "Lab 5: Relationships in Bechdel"
author: "Your Name Here"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
#Include all libraries here



#Import your data here 



```

## Task 1: Compute scores for scales

Be sure to compute the appropriate total scores for the ESS, ASHS and Attention variables.

Once completed, only include students who have "passed" the attention check (i.e., having a score of 4 or higher on the attention sum total).

*hint: use dplyr and filter*

```{r}
## Task 1: Computing scales



```

## Task 2: Roommates and Daytime Sleepiness (A Group Comparison)

**The Research Question:** "Do students who have a roommate report different levels of daytime sleepiness compared to students who do not have a roommate?"

**Your Steps:**

1.  **Identify & Model:**

    -   What is the predictor (IV) and what is the outcome (DV)?

    -   Are they categorical or continuous?

    -   Based on this, what is the appropriate model family? Write out the model in R formula syntax (`outcome ~ predictor`).

2.  **Describe & Visualize:**

    -   Calculate the mean and standard deviation of `ess_score` for both groups (those with and without a roommate).

    -   Create a boxplot to visualize the distribution of `ess_score` for each group. Make sure your plot is clearly labeled.

```{r}
## Describe your data. Look back to past labs where there was a group_by() and then summarize() type of function




## We also created some boxplots before


```

3.  **Analyze:**

    -   Run the appropriate statistical test in R to determine if there is a significant difference between the groups.

    ```{r}
    ## Conduct your analysis


    ```

4.  **Interpret & Conclude:**

    -   What is the p-value from your test?

    -   Based on the test and your descriptive statistics, write a one-sentence conclusion that directly answers the research question.

    -   **Critical Thinking:** This is an observational study. Can you conclude that having a roommate *causes* a change in sleepiness? Why or why not?

## Bonus: Sleep Habits and Age (An Association)

This question is not required. But you will get some bonus points if you decide to do it!

**The Research Question:** "Is there an association between a student's sleep habits and their age?"

**Your Steps:**

1.  **Identify & Model:**

    -   What is the predictor (IV) and what is the outcome (DV)?

    -   Are they categorical or continuous?

    -   Based on this, what is the appropriate model family? Write out the model in R formula syntax.

2.  **Visualize:**

    -   Create a scatterplot to visualize the relationship between `ashs_score` (sleep habits) and `age`.

    -   Add a line of best fit to the plot. Make sure your plot is clearly labeled.

3.  **Analyze:**

    -   Run the appropriate statistical test in R to determine if there is a significant association between the two variables.

4.  **Interpret & Conclude:**

    -   What is the correlation coefficient (`r`) and the p-value from your test?

    -   Based on these results, write a one-sentence conclusion that describes the nature and significance of the relationship.

*End of Lab. Don't forget to Knit! 🧶*
