---
title: "Lab 3: Exploring Bechdel Test Movie Data"
author: "Your Name Here"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Scenario and Goal

In this lab, you will act as a data journalist exploring a dataset on movies. We will use the data from the FiveThirtyEight story “The Dollar-And-Cents Case Against Hollywood’s Exclusion of Women.”

This analysis is about the [Bechdel test](https://en.wikipedia.org/wiki/Bechdel_test), a measure of the representation of women in fiction.

Your goal is to **import, describe, and visualize** this data to understand the characteristics of movies in the dataset and see if there are relationships between a movie's budget, its box office gross, and its Bechdel Test rating. This is the critical first step in any analysis, known as **Exploratory Data Analysis (EDA)**.

------------------------------------------------------------------------

### Exercise 1: Importing and Inspecting

First, you need to set up your RMarkdown to get it ready for importing the data and using the appropriate libraries. Be sure to have all libraries listed here in the first code chunk along with importing the data. I should not see any lines that say `install.packages()`.

I will attempt to reproduce your output in my own computer, so be sure that your code is reproducible.

```{r ex1}
# Load the libraries


# Import the dataset 


# Get an overview of your data

```

**Question 1:** Look at the output from your overview. How many total movies are in this database? And what year is the latest movie?

**Your Answer:** [Type your answer here]

**Question 2:** Calculate the Average budget of the whole dataset. Then, calculate the average for only movies in the year 2000.

**Your Answer:** [Type your answer here]

------------------------------------------------------------------------

### Exercise 2: Grouped Descriptive Statistics

Averages for the whole dataset are useful, but we are often more interested in comparing averages *between groups*. Let's see if the average budget differs for movies that pass the Bechdel Test versus those that fail. The `binary` variable tells us this ("PASS" or "FAIL").

```{r ex2}
# Task: Use group_by() and summarise() to calculate the mean and standard deviation
# of `budget` for each category in the `binary` variable.
# Create a new object called `budget_summary`.

# Hint: The syntax is: dataset %>% group_by(grouping_variable) %>% summarise(new_mean_name = mean(variable), new_sd_name = sd(variable))



# Print your new `budget_summary` object to see the results.
budget_summary
```

**Question 3:** Based on your summary table, do movies that pass or fail the Bechdel test have a higher average (mean) budget?

**Your Answer:** [Type your answer here]

------------------------------------------------------------------------

### Exercise 3: Visualizing a Distribution with a Histogram

Let's visualize the distribution of movies based on their IMDB rating (`imdb_rating`). A histogram is the perfect tool for this, as it shows the frequency of ratings in different bins.

```{r ex3}
# Task: Create a histogram of the `imdb_rating` variable from the main `movies_data` dataset.
# Add a title and more descriptive axis labels.

# Write your ggplot code here:
ggplot(data = movies_data, aes(x = imdb_rating)) +
  geom_histogram(bins = 20, color = "white", fill = "darkred") +
  labs(title = "Distribution of IMDB Ratings for Movies",
       x = "IMDB Rating",
       y = "Number of Movies")
```

**Question 4:** Describe the shape of the distribution you see in the histogram. Is it symmetric (like a bell curve), or is it skewed in one direction? Where do most movies' ratings seem to be clustered?

**Your Answer:** [Type your answer here]

------------------------------------------------------------------------

### Exercise 4: Comparing Groups with a Boxplot

Now let's visually compare the inflation-adjusted domestic gross (`domgross_2013`) for movies that pass the Bechdel test versus those that fail. A boxplot is an excellent way to see differences in the median and spread between groups.

```{r ex4}
# Task: Create a boxplot to compare `domgross_2013` for movies that "PASS" vs. "FAIL".
# Put `binary` on the x-axis and `domgross_2013` on the y-axis.

# Write your ggplot code here:
ggplot(data = movies_data, aes(x = binary, y = domgross_2013)) +
  geom_boxplot() +
  labs(title = "Domestic Gross by Bechdel Test Result",
       x = "Bechdel Test Result",
       y = "Domestic Gross (2013 Dollars)")
```

**Question 5:** Look at the boxplot. The thick horizontal line in the middle of each box represents the median. Does there appear to be a large difference in the median domestic gross between movies that pass and fail the test?

**Your Answer:** [Type your answer here]

------------------------------------------------------------------------

*End of Lab 3. You've now practiced the complete exploratory data analysis workflow!*
