---
title: "Unit 5 - Descriptives"
subtitle: "PSYC 640 - Fall 2024"
author: "Dustin Haraden, PhD"
format: 
  revealjs:
    multiplex: true
    scrollable: true
    slide-number: true
    incremental: false
    touch: true
    code-overflow: wrap
    theme: dark
execute: 
  echo: true
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r, include = F}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Reminders

::: nonincremental
-   Journal Entries

-   Lab 1 to be posted soon (if we have time today I will try to introduce it)
:::

------------------------------------------------------------------------

## Last Class

::: nonincremental
-   Used various functions to wrangle data
    -   `select()`
    -   `filter()`
    -   `mutate()`
:::

------------------------------------------------------------------------

## Today...

Entering into the world of describing our data with numerical values.

```{r, results = 'hide', message = F, warning = F}
# File management
library(here)
# for dplyr, ggplot2
library(tidyverse)
#Loading data
library(rio)
# for descriptives
library(psych)

#Remove Scientific Notation 
options(scipen=999)
```

# Let's get caught up

------------------------------------------------------------------------

Getting the code a little cleaner. And doing it all in one step

```{r}
#import sleep data - Your path may be different
sleep_data <- import(here("lectures", "data", "Sleep_Data.csv")) %>% 
  
  #Select only the variables we are interested in
  select(-c('StartDate', 'EndDate', 'Status', 'Progress', 
  'Duration__in_seconds_', 'Finished', 'RecordedDate',
  'DistributionChannel', 'UserLanguage', 'SC0', 'SC1',
  'SC2', 'SC3', 'SC4', 'SC5', 'Attention')) %>% 
  
  #rename the variable names so they make sense
  setNames(c('age', 'gender', 'roommate', 'other_sleep', 'bed_read', 'bed_study', 'bed_hw', 'attention1', 'bed_internet',
                       'bed_tv', 'bed_eat', 'bed_friends', 'bed_videogame', 'bed_readp', 'BL_1', 'BL_2', 'BL_3', 'BL_4',
                       'BL_5', 'attention2', 'BL_6', 'BL_7', 'sleepsat1', 'sleepsat2', 'sleepsat3', 'sleepsat4', 'sleepsat5',
                       'sleepsat6', 'ESS1', 'ESS2', 'ESS3', 'ESS4', 'ESS5', 'ESS6', 'ESS7', 'ESS8', 'ESS00', 'ashs1', 'ashs2', 
                       'ashs3', 'ashs4', 'ashs5', 'attention3', 'ashs6', 'ashs7', 'ashs8', 'ashs9', 'ashs10', 'ashs11', 
                       'ashs12', 'ashs13', 'ashs14', 'ashs15', 'ashs16', 'ashs17', 'attention4', 'ashs18', 'ashs19', 
                       'ashs20', 'ashs21', 'ashs22', 'ashs23', 'ashs24', 'ashs26', 'ashs27', 'ashs28', 'attention5', 
                       'ashs29', 'ashs30', 'ashs31', 'ashs32', 'ashs33', 'id')) %>% 
  
  # Only include college aged participants
  filter(age >= 17, age <= 25) %>% 
  
  # Rescale the ESS
  mutate(ESS1m1 = ESS1 - 1, 
         ESS2m1 = ESS2 - 1,
         ESS3m1 = ESS3 - 1,
         ESS4m1 = ESS4 - 1,
         ESS5m1 = ESS5 - 1,
         ESS6m1 = ESS6 - 1,
         ESS7m1 = ESS7 - 1, 
         ESS8m1 = ESS8 - 1) %>% 
  
  # Create a total score for the ESS
  mutate(ess_total = ESS1m1 + ESS2m1 + ESS3m1 + 
           ESS4m1 + ESS5m1 + ESS6m1 + ESS7m1 + ESS8m1) %>% 
  
  # Recode the attention check variables
  mutate(attention1r = recode(attention1, 
                              `1`=0,
                              `2`=0,
                              `3`=0,
                              `4`=0,
                              `5`=1),
         attention2r = recode(attention2, 
                              `1`=1,
                              `2`=0,
                              `3`=0,
                              `4`=0,
                              `5`=0),
         attention3r = recode(attention3, 
                              `1`=0,
                              `2`=0,
                              `3`=0,
                              `4`=1,
                              `5`=0),
         attention4r = recode(attention4, 
                              `1`=1,
                              `2`=0,
                              `3`=0,
                              `4`=0,
                              `5`=0,
                              `6`=0),
         attention5r = recode(attention5, 
                              `1`=1,
                              `2`=0,
                              `3`=0,
                              `4`=0,
                              `5`=0,
                              `6`=0)) %>% 
  
  # Compute a total Attention Score
  mutate(att_sum = (attention1r + attention2r + attention3r +
                      attention4r + attention5r))
```

------------------------------------------------------------------------

## Address the Inattention

Create two new datasets labeled (1) “data_attend” and (2) “data_distract”. In each dataset have those who were paying attention in the “data_attend” and those who were not in the “data_distract”. Paying attention is operationalized as having a score of 5 on the aggregated variable.

Get the sample size of each of these datasets (use Google to search for things like number of rows)

------------------------------------------------------------------------

## Save the cleaned up file

```{r eval = FALSE}
write.csv(sleep_data, 
          here("lectures", "data", "named_Sleep_Data.csv"), 
          row.names = FALSE)

```

This is now a file that we will use in the future to import. That way we don't have to make sure that the code above is in all of our scripts/recipes moving forward!

# Descriptive Statistics

------------------------------------------------------------------------

## Measures of Central Tendency

For a given set of observations, measures of central tendency allow us to get the “gist” of the data.

They tell us about where the “average” or the “mid-point” of the data lies or how much deviation there is from a central point.

Let’s take a look at the data that we have already loaded in, and complete some of these tasks.

------------------------------------------------------------------------

### Mean

A quick way to find the mean is to use the aptly named `mean()` function from base R. Use this function to get the average `age` and `ess total` in our dataset.

```{r}
mean(sleep_data$age)

mean(sleep_data$ess_total)
```

$\bar{X} = \frac{X_1+X_2+...+X_{N-1}X_N}{N}$

Oh no! We forgot to account for the missing variables in our variable! We got NA! The reason for this is that the mean is calculated by using every value for a given variable, so if you don’t remove (or impute) the missing values before getting the mean, it won’t work.

Let’s try that again, but using the additional argument to eliminate (or remove) the NA’s from the variable prior to computing the mean. 

|                                         |
|-----------------------------------------|
| **mean(sleep_data\$age, na.rm = TRUE)** |

### Median

The median is the middle value of a set of observations: 50% of the data points fall below the median, and 50% fall above.

To find the median, we can use the median() function. Use it on the age variable.

## Measures of Variability

### Range

The range gives us the distance between the smallest and largest value in a dataset. You can find the range using the range() function, which will output the minimum and maximum values. Find the range of the age variable.

### Variance and Standard Deviation

To find the variance and standard deviation, we use var() and sd(), respectively. Find the variance and standard deviation of the age variable.

# Summarizing Data

So far we have been calculating various descriptive statistics (somewhat painstakingly) using an assortment of different functions. So what if we have a dataset with a bunch of variables we want descriptive statistics for? Surely we don’t want to calculate descriptives for each variable by hand…

Fortunately for us, there is a function called describe() from the {psych} package, which we can use to quickly summarize a whole set of variables in a dataset.

Be sure to first install the package prior to putting it into your library code chunk. Reminder: anytime you add a library, be sure you actually run the code line **library(psych)**. Otherwise, you will have a hard time trying to use the next functions. 

Let’s use it with our sleep dataset!

## describe()

This function automatically calculates all of the descriptives we reviewed above (and more!). Use the describe() function from the **psych** package on the entire sleep_data dataset.

**Notes:** If you load a library at the beginning, you can directly call any function from it. Instead, you can call a function by library_name::function_name without loading the entire library.

+--------------------------------------------------+
| **psych::describe(sleep_data)**                  |
|                                                  |
| **\# or if you have already loaded the library** |
|                                                  |
| **describe(sleep_data)**                         |
+--------------------------------------------------+

NOTE: Some variables are not numeric and are **categorical** variables of type character. By default, the describe() function forces non-numeric variables to be numeric and attempts to calculate descriptives for them. These variables are marked with an asterisk (\*). In this case, it doesn’t make sense to calculate descriptive statistics for these variables, so we get a warning message and a bunch of NaN’s and NA’s for these variables.

A better approach would be to remove non-numeric variables before you attempt to run numerical calculations on your dataset.
